---
title: "HT switch analysis: T_HT and H_HT"
author: "Project"
date: last-modified
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
    fig-align: center
---

## Scope

This report focuses on the HT reference transcriptome datasets:

- `T_HT` (anchor)
- `H_HT` (supplementary)

Ranking is performed with q-value cutoffs, ordered by `max_abs_dif_switching`, and flags genes with two or more significant isoforms that switch in opposing directions.

```{r}
#| label: setup
#| message: false
library(knitr)
library(dplyr)

source("_setup.R")
fig <- make_fig("ht_top_switch_plots")
```

```{r}
#| output: asis
cat(reduction_caveat())
```

```{r}
#| label: load-data
#| message: false
require_outputs(
  c(
    "top_switching_genes_T_HT_anchor.csv",
    "top_switching_genes_H_HT_supplementary.csv",
    "isoform_overlap_T_HT_vs_H_HT_summary.csv",
    "isoform_overlap_T_significant_in_H_context.csv"
  ),
  "Missing HT analysis outputs. Run: Rscript scripts/06_visualization.R"
)

t_top <- read_tab("top_switching_genes_T_HT_anchor.csv")
h_top <- read_tab("top_switching_genes_H_HT_supplementary.csv")
ov_sum <- read_tab("isoform_overlap_T_HT_vs_H_HT_summary.csv")
ov_detail <- read_tab("isoform_overlap_T_significant_in_H_context.csv")
```

## T_HT panel

### Top 30 genes (ranked by max_abs_dif_switching)

```{r}
knitr::kable(
  t_top,
  digits = 4,
  caption = "T_HT top 30 genes (q-cutoff filtered; ranked by max_abs_dif_switching)"
)
```

### T_HT genes with multi-isoform opposing-direction switches

```{r}
t_interest <- t_top |>
  filter(.data$has_multi_isoform_opposing_switch %in% TRUE) |>
  arrange(desc(.data$max_abs_dif_switching), desc(.data$n_switching_isoforms))
if (nrow(t_interest) == 0) {
  knitr::asis_output("_No T_HT top-30 genes met the multi-isoform opposing-direction criterion._")
} else {
  knitr::kable(
    t_interest,
    digits = 4,
    caption = "T_HT genes of particular interest (>=2 significant isoforms changing in opposing directions)"
  )
}
```

### T_HT switch plots

```{r}
# include_graphics() must be called ONCE on a vector. Looping over files and calling
# it per iteration returns a value that is never printed, which is why this report
# previously rendered with zero images despite 60 switchPlots existing on disk.
t_figs <- sort(list.files(
  fig(),
  pattern = "^switch_plot_T_HT_anchor_rank[0-9]{2}_.*\\.png$",
  full.names = TRUE
))
if (!length(t_figs)) {
  knitr::asis_output("_No T_HT switch plots found._")
} else {
  knitr::include_graphics(utils::head(t_figs, 12))  # first 12 keeps the report navigable
}
```

## H_HT panel

### Top 30 genes (ranked by max_abs_dif_switching)

```{r}
knitr::kable(
  h_top,
  digits = 4,
  caption = "H_HT top 30 genes (q-cutoff filtered; ranked by max_abs_dif_switching)"
)
```

### H_HT genes with multi-isoform opposing-direction switches

```{r}
h_interest <- h_top |>
  filter(.data$has_multi_isoform_opposing_switch %in% TRUE) |>
  arrange(desc(.data$max_abs_dif_switching), desc(.data$n_switching_isoforms))
if (nrow(h_interest) == 0) {
  knitr::asis_output("_No H_HT top-30 genes met the multi-isoform opposing-direction criterion._")
} else {
  knitr::kable(
    h_interest,
    digits = 4,
    caption = "H_HT genes of particular interest (>=2 significant isoforms changing in opposing directions)"
  )
}
```

### H_HT switch plots

```{r}
h_figs <- sort(list.files(
  fig(),
  pattern = "^switch_plot_H_HT_supplementary_rank[0-9]{2}_.*\\.png$",
  full.names = TRUE
))
if (!length(h_figs)) {
  knitr::asis_output("_No H_HT switch plots found._")
} else {
  knitr::include_graphics(utils::head(h_figs, 12))
}
```

## T vs H overlap panel

This panel goes beyond "top/significant only" and asks:

- For isoforms significant in `T_HT`, are they also detected in `H_HT`?
- If detected in `H_HT`, are they also significant there?
- If not significant in `H_HT`, what is the observed direction/magnitude context?

### Overlap summary

```{r}
knitr::kable(ov_sum, digits = 4, caption = "Isoform overlap summary: T_HT vs H_HT")
```

### T-significant isoforms in H context (including non-significant H)

```{r}
ov_show <- ov_detail |>
  arrange(desc(abs(.data$T_dIF_n)), .data$T_q_i) |>
  select(
    gene_id, gene_name, isoform_id,
    T_dIF_n, T_q_i, T_q_g, T_significant,
    H_dIF_n, H_q_i, H_q_g, H_significant,
    same_direction, overlap_note
  )

knitr::kable(
  utils::head(ov_show, 100),
  digits = 4,
  caption = "Top T-significant isoforms mapped to H context (first 100 rows)"
)
```
