---
title: "HT switch analysis: T_HT and H_HT"
author: "Project"
date: last-modified
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
    fig-align: center
---

## Scope

This report focuses on the HT reference transcriptome datasets:

- `T_HT` (anchor)
- `H_HT` (supplementary)

Ranking is performed with q-value cutoffs, ordered by `max_abs_dif_switching`, and flags genes with two or more significant isoforms that switch in opposing directions.

```{r}
#| label: setup
#| message: false
library(knitr)
library(dplyr)

.infer_project_root <- function() {
  p <- getwd()
  for (i in 1:20) {
    if (file.exists(file.path(p, "config", "config.yml"))) {
      return(normalizePath(p, mustWork = TRUE))
    }
    p2 <- dirname(p)
    if (identical(p, p2)) break
    p <- p2
  }
  stop("Could not find config/config.yml")
}
knitr::opts_knit$set(root.dir = .infer_project_root())
```

```{r}
#| label: load-data
#| message: false
t_path <- "results/tables/top_switching_genes_T_HT_anchor.csv"
h_path <- "results/tables/top_switching_genes_H_HT_supplementary.csv"
ov_sum_path <- "results/tables/isoform_overlap_T_HT_vs_H_HT_summary.csv"
ov_detail_path <- "results/tables/isoform_overlap_T_significant_in_H_context.csv"

if (!file.exists(t_path) || !file.exists(h_path) || !file.exists(ov_sum_path) || !file.exists(ov_detail_path)) {
  stop("Missing HT analysis outputs. Run: Rscript scripts/06_visualization.R")
}

t_top <- read.csv(t_path, stringsAsFactors = FALSE, check.names = FALSE)
h_top <- read.csv(h_path, stringsAsFactors = FALSE, check.names = FALSE)
ov_sum <- read.csv(ov_sum_path, stringsAsFactors = FALSE, check.names = FALSE)
ov_detail <- read.csv(ov_detail_path, stringsAsFactors = FALSE, check.names = FALSE)
```

## T_HT panel

### Top 30 genes (ranked by max_abs_dif_switching)

```{r}
knitr::kable(
  t_top,
  digits = 4,
  caption = "T_HT top 30 genes (q-cutoff filtered; ranked by max_abs_dif_switching)"
)
```

### T_HT genes with multi-isoform opposing-direction switches

```{r}
t_interest <- t_top |>
  filter(.data$has_multi_isoform_opposing_switch %in% TRUE) |>
  arrange(desc(.data$max_abs_dif_switching), desc(.data$n_switching_isoforms))
if (nrow(t_interest) == 0) {
  knitr::asis_output("_No T_HT top-30 genes met the multi-isoform opposing-direction criterion._")
} else {
  knitr::kable(
    t_interest,
    digits = 4,
    caption = "T_HT genes of particular interest (>=2 significant isoforms changing in opposing directions)"
  )
}
```

### T_HT switch plots

```{r}
t_figs <- list.files(
  "results/figures/ht_top_switch_plots",
  pattern = "^switch_plot_T_HT_anchor_rank[0-9]{2}_.*\\.png$",
  full.names = TRUE
)
t_figs <- sort(t_figs)
if (!length(t_figs)) {
  knitr::asis_output("_No T_HT switch plots found._")
} else {
  # Render first 12 to keep report navigable.
  for (f in utils::head(t_figs, 12)) {
    knitr::include_graphics(f)
    cat("\n\n")
  }
}
```

## H_HT panel

### Top 30 genes (ranked by max_abs_dif_switching)

```{r}
knitr::kable(
  h_top,
  digits = 4,
  caption = "H_HT top 30 genes (q-cutoff filtered; ranked by max_abs_dif_switching)"
)
```

### H_HT genes with multi-isoform opposing-direction switches

```{r}
h_interest <- h_top |>
  filter(.data$has_multi_isoform_opposing_switch %in% TRUE) |>
  arrange(desc(.data$max_abs_dif_switching), desc(.data$n_switching_isoforms))
if (nrow(h_interest) == 0) {
  knitr::asis_output("_No H_HT top-30 genes met the multi-isoform opposing-direction criterion._")
} else {
  knitr::kable(
    h_interest,
    digits = 4,
    caption = "H_HT genes of particular interest (>=2 significant isoforms changing in opposing directions)"
  )
}
```

### H_HT switch plots

```{r}
h_figs <- list.files(
  "results/figures/ht_top_switch_plots",
  pattern = "^switch_plot_H_HT_supplementary_rank[0-9]{2}_.*\\.png$",
  full.names = TRUE
)
h_figs <- sort(h_figs)
if (!length(h_figs)) {
  knitr::asis_output("_No H_HT switch plots found._")
} else {
  # Render first 12 to keep report navigable.
  for (f in utils::head(h_figs, 12)) {
    knitr::include_graphics(f)
    cat("\n\n")
  }
}
```

## T vs H overlap panel

This panel goes beyond "top/significant only" and asks:

- For isoforms significant in `T_HT`, are they also detected in `H_HT`?
- If detected in `H_HT`, are they also significant there?
- If not significant in `H_HT`, what is the observed direction/magnitude context?

### Overlap summary

```{r}
knitr::kable(ov_sum, digits = 4, caption = "Isoform overlap summary: T_HT vs H_HT")
```

### T-significant isoforms in H context (including non-significant H)

```{r}
ov_show <- ov_detail |>
  arrange(desc(abs(.data$T_dIF_n)), .data$T_q_i) |>
  select(
    gene_id, gene_name, isoform_id,
    T_dIF_n, T_q_i, T_q_g, T_significant,
    H_dIF_n, H_q_i, H_q_g, H_significant,
    same_direction, overlap_note
  )

knitr::kable(
  utils::head(ov_show, 100),
  digits = 4,
  caption = "Top T-significant isoforms mapped to H context (first 100 rows)"
)
```
